A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2020-03-27 paper/paper - revision.Rnw
Flesh out introduction, make formal to-do list

NEW OUTLINE

In the United States, forensic analysis of shoe print evidence in criminal cases is varied and difficult. The process of collecting print evidence from a crime scene is not standardized, and it is often impossible to conduct an individualizing analysis when a print is partial or of low quality. In cases where individualizing characteristics are not considered (estimated at 95\% of cases in the United States according to some experts \footnote{Leslie Hammer, presentation to CSAFE on March 5, 2018}), it is useful to be able to assess the random match probability. In other words, given a questioned shoe print, what is the probability that a randomly selected shoe in the relevant population has a similar tread pattern? Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns (e.g., knockoffs, same models over different years, etc)

From a statistical perspective, random match probability could be calculated by sampling a large number of shoes from a given population and counting the number of shoes in the sample whose tread patterns are similar to the given print. In practice, one potential method of sampling could be a low-profile scanner placed in a high traffic area. Such a scanner could take pictures of shoes walking past, and those images could then be used to create a database of shoes relevant to the local area. \mt{Are cameras more robust/weatherproof than pressure scanners? Why images instead of EverOS impressions?} \mt{(Note: a grant proposal for such a scanner has been approved by NIJ, and there currently exists a working prototype.)} Assuming that such a database of images is available, the task becomes assessing the similarity of the shoe tread patterns in the images. This is done by identifying a set of relevant features and using image analysis methods to detect the presence or absence of the features in each image, and then defining similarity measures based on those features to determine how many shoes have a tread pattern similar to the pattern of interest.

CNNS have become the gold standard of image recognition. \mt{salvage the other intro to explain why simpler methods aren't sufficient.} There is precedent for using machine learning to identify tread patterns \mt{hunt through papers to find relevant references}

>>>>

\mt{General paper to-do:}
\mt{1. Find a paper to cite the 95\% statistic}
\mt{Option: Shoe marks are most common at a crime scene (Rida et al., 2019)}
\mt{2. A zotero reference is missing details, remember to go find/fix it}
\mt{3. Cite papers with manual data collection}
\mt{4. Susan will send pictures of the prototype scanner from the grant proposal}
\mt{5. Argue in methods that we know real image quality will be lower but we needed to start somewhere}
\mt{6. Our goal is to narrow down what type of model works and assess the model(s) limitations}
\mt{7. Add to conclusion: Overall, this concept seems to work, but we need to modify the classificiation scheme. Thus, this work is publishable, because it needs work and community input. Need more research into human comphrehensible and yet useful classification schemes for this purpose. We need better class scheme, important and promising, getting this right matters}
\mt{8. Update abstract to reflect new framing of paper}
\mt{}

In the United States, forensic analysis of shoe print evidence in criminal cases is varied and difficult. The process of collecting print evidence from a crime scene is not standardized, and it is often impossible to conduct an individualizing analysis when a print is partial or of low quality. In fact, some experts estimate that 95\% of cases in the United States do not consider individualizing characteristics of shoeprint evidence \fix{(cite)}. \mt{(see list above)}
%\footnote{Leslie Hammer, presentation to CSAFE on March 5, 2018}).
Even so, in these cases it still may be \mt{both possible and useful} to assess the random match probability of the footwear evidence \mt{at the level of class characteristics}. In other words, given a questioned shoe print, what is the probability that a randomly selected shoe in the relevant population has a similar tread pattern? Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns (e.g., knockoffs, similar models over different years, etc).

From a statistical perspective, random match probability could theoretically be calculated by sampling a large number of shoes from a given population and counting the number of shoes in the sample whose tread patterns are similar to the given print. In practice, one potential method of sampling could be a low-profile scanner placed in a high traffic area. Such a scanner could take \mt{impressions or} pictures of shoes walking past, and those scans could then be used to create a database of shoes relevant to the local area.

However, studies by \fix{authors (yr)} and \fix{authors (yr)} demonstrate that even \fix{[a couple hundred]} shoe prints constitute an unweildy amount of data requiring \fix{[a large amount of data processing]}. Thus, in parallel with the development of such a scanner, it is necessary to develop a way to process large amounts of data automatically. \mt{The primary focus of this paper is to develop one such method for large-scale automatic data processing; in this case, "processing" refers to assessing} the similarity of the shoe tread patterns in the collected data. %\fix{Cut this?: This is done by identifying a set of relevant features and using image analysis methods to detect the presence or absence of the features in each image, and then defining similarity measures based on those features to determine how many shoes have a tread pattern similar to the pattern of interest.}

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, early prototypes of a footwear scanner have suggested that cameras are more robust than pressure scanners for placement in public locations that may experience inclement weather or other non-controlled conditions. Thus, \fix{this is an image recognition problem}.

% From original intro:
Several attempts have been made to automatically identify features in both prints and shoe tread images, with low-level image analysis methods including fractal decomposition \citep{alexanderAutomaticClassificationRecognition1999}, scale-invariant feature recognition \citep{pavlouAutomaticExtractionClassification2006}, Fourier-Mellin transformation \citep{guehamAutomaticRecognitionShoeprints2008}, and other classical image analysis methods. Some of these low-level methods perform relatively well in good conditions \citep{luostarinenMeasuringAccuracyAutomatic2014a} but have degraded performance under conditions which are suboptimal, including those commonly found at crime scenes.

% From original intro:
In the wider field of computer vision, attention has turned to more robust methods for image recognition, such as convolutional neural networks (CNNs), which are capable of achieving near-human accuracy under even degraded image conditions \citep{krizhevskyImageNetClassificationDeep2012}. These networks are designed to mimic the process of human vision, and typically involve application of sets of filters, some of which mimic the filters used in the low-level techniques used in early automatic classification attempts. CNNs have been applied in footwear forensics, making use of more general neural networks optimized to detect objects found in natural scenes, such as trees, animals, buildings, and cars; the pre-trained network's features are then used to compare shoes or prints to determine how well they match \citep{kongCrossDomainImageMatching2019a,kongCrossDomainForensicShoeprint2017,zhangAdaptingConvolutionalNeural2017}. Neural network methods appear to be more successful than traditional image analysis techniques, but the features used for matching are not generally informative for humans, who typically classify shoes by patterns and spatial relationships \citep{davisIntelligenceApproachFootwear1981,grossVariabilitySignificanceClass2013}.

There is precedent for using machine learning to identify tread patterns \mt{hunt through papers to find relevant references}

% ORIGINAL INTRO:
